---
title: "An Overview of TidyGEO and Some Helpful Resources"
author: "Avery Bell"
date: "June 25, 2019"
output: html_document
---


## Table of Contents {#contents}
1. __[Introduction to TidyGEO](#intro-to-tidygeo)__
   + [Introduction to Gene Expression Omnibus](#intro-to-geo)
   + [Motivation for the app](#app-motivation)
2. __[Use Case](#use-case)__
   + [Cleaning up a GEO dataset](#geo-cleaning-steps)
   + [Possible application](#example)
3. __[Shiny Apps](#shiny-apps)__
   + [Basic format](#shiny-apps-overview)
   + [Resources](#shiny-apps-resources)
4. __[TidyGEO Code](#tidygeo-code)__
   + [Directory structure](#directory-structure)
   + [File contents](#file-contents)
   + [Style guide](#style-guide)
   + [Functions and how to use them](#tidygeo-functions)
5. __[Next Steps](#next-steps)__
   + [Known bugs](#known-bugs)
   + [TODO items](#todo-items)
   + [Possible future features](#future-features)
5. __[Miscellaneous](#misc)__
   + [Error log](#error-log)
   + [Shiny/R tips and observations](#observations)
   + [Tips for time testing](#time-testing)
   + [Other resources](#other-resources)
       - [On RMarkdown](#Rmarkdown-resource)

------------------------

## Introduction {#intro-to-tidygeo}

### Gene Expression Omnibus {#intro-to-geo}
An online repository of array- and sequence-based data. Scientists can go here for publicly available data to look at genes, proteins, etc. that interest them. One of the largest databases of its kind. Submissions are [MIAME](https://www.ncbi.nlm.nih.gov/geo/info/MIAME.html) compliant but the metadata associated with the submissions (i.e., descriptions of the sample sources, such as patient age and disease assessment) are not regulated, so they are often messy and hard to analyse using computational tools.

GEO data can be accessed via an R package called GEOQuery, but scientists with no R coding experience must learn the language in order to use this tool, and manual data extraction can often be tedious and error-prone, especially with large batches of datasets that are necessary for certain computational tools to perform their analysis.

There are some tools that have attempted to clean up the metadata and to enable scientists to access it, but these tools either perform the analysis for the user (not giving them access to the cleaned-up data) or rely on crowd-curating efforts, which are not able to keep up with the growing number of datasets on GEO.

### TidyGEO {#app-motivation}
TidyGEO is an app that lets scientists (even those with little coding experience) download and clean up GEO data to use for their own purposes.

[&uarr; To top](#contents)

------------------------

## Use case {#use-case}
[GSE112914](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE112914): Differential gene expression analysis of MYCN-amplified neuroblastoma cells after Doxycycline inducible shRNA knockdown of JMJD6 gene expression

### [BRB Array Tools](https://brb.nci.nih.gov/BRB-ArrayTools/)
A package that helps visualize and perform statistical analysis on Microarray gene expression, copy number, methylation and RNA-Seq data.

### Cleaning up the data {#geo-cleaning-steps}
[TidyGEO](https://tidygeo.shinyapps.io/tidygeo/)

1. Filter out unnecessary columns (to be less overwhelming)
2. Split key-value pairs
3. Split multiple values in a column
3. Shift cells
4. Substitute doxy and control in replicate for NA
5. Substitute "repeat" for ""
6. Filter out NAs in replicate, or maybe duplicates in replicate
7. Download clinical data
8. Briefly look at assay and feature data

### Analysis {#example}
Now you can look at the effect of replicates on the expression measurements.

[&uarr; To top](#contents)

------------------------

## Shiny Apps {#shiny-apps}
Shiny is an R package that makes it easy to build interactive web apps straight from R.

### The basics {#shiny-apps-overview}
* [Structure of a Shiny App](https://shiny.rstudio.com/tutorial/written-tutorial/lesson1/)
* UI
    + The UI lays out what the app looks like. It is made up of [widgets](https://shiny.rstudio.com/gallery/widget-gallery.html) that allow the user to interact with the app and [layouts](https://shiny.rstudio.com/articles/layout-guide.html) that define how the widgets are arranged on the page.
* Server
    + The server is a function made up of
        - Outputs that define how UI objects should behave (e.g., renderText, renderTable, etc.)
        - Listeners that react to a user's input (e.g., observeEvent)
    + Basically, it does the __work__ of the app to actually make it interactive
* Running an app
    + Call to shinyApp function
    + Must have either `app.R` with UI and server function or `ui.R` and `server.R`
    + Builds the app in the directory you name or the working directory, if no name is given
* Other resources
    + reactive values
        - the difference between reactive values and global variables
    + www/ directory
        - must be named www/, recognized by the app as a place where resources (graphics, CSS, JS scripts) are kept
    + CSS file
        - Cascading Style Sheets: an HTML file that defines how the app should look
        - Shiny automatically uses Bootstrap styling but you can customize this in the CSS file
        - generally it is good practice to use a CSS file rather than defining colors, sizes, etc. inline
    + JS scripts
        - for more complicated reactivity, Shiny may not be flexible enough to meet your needs
        - you can write JavaScript that will define how certain objects react to button clicks, etc.
        - this can also be defined inline but it is good practice to put this in a separate file
    + helper functions
        - code that is run in multiple parts of the app can be put in a function to reduce repetition
        - often used by the server function to do work, but also may save some space in the UI part
        - these are often kept in a separate file to clean up code
    + graphics
        - kept in the www/ directory
        - may include browser icon, logos, icons, etc.

### Resources {#shiny-apps-resources}
* [Introduction to Shiny video](https://www.rstudio.com/resources/webinars/introduction-to-shiny/)
* [Shiny Video and Written Tutorials](https://shiny.rstudio.com/tutorial/)
* [R Style Guide](https://style.tidyverse.org/)

[&uarr; To top](#contents)

------------------------

## TidyGEO Code {#tidygeo-code}

### TidyGEO structure {#directory-structure}
* Root
    + `.gitignore`
        - Lists files that should not be updated on GitHub
    + `Code_Review_Notes.Rmd`
        - Some notes from my Code Review--pretty much the same as this document except less detailed
    + `geocurate_repo.Rproj`
        - Some settings for the old Rproject (an RStudio thing)
    + `README.md`
        - The readme for the GitHub repo
    + `TidyGEO/`
        - Contains all the code
    + `TidyGEO_Directory_Guide.Rmd`
       - This file
* Level 1 - TidyGEO
    + `app.R`
        - sources all the R scripts from the ui and server directories
        - creates the ui object and the server function
        - creates some lists of reactiveValues that are used in the app
        - contains some instructions for how column navigation should work
        - contains shinyApp function to knit UI and server
    + `deploy_app.R`
        - updates the list of available GEO series to choose from
        - updates the list of formatting functions that can be included in the scripts the user can download
        - launches the app to shinyapps.io
    + `error_log.md`
        - a list of errors I have encountered and how I fixed them
    + `file_manifest.txt`
        - defines which files are necessary when publishing the app to the web
    + `generate_profile.R`
        - generates a profile for the app, see the [time testing tips](#time-testing) section
    + `generate_rscript_functions.R`
        - generates a list of functions that can be used in the scripts the user can download
    + `tidygeo-install.R`
        - contains the packages needed to run the app and use any of the functions in the app (some packages may be             extraneous--this hasn't been updated in a while)
    + `tidygeo_functions.R`
        - some functions for the app that don't really fall under the "ui", "server", or "variables" categories;                mostly for script writing
    + `tidygeo_variables.R`
        - consists of two parts: global variables and naming conventions
        - global variables are constants used by multiple parts of the app
        - naming conventions are patterns for which variables in the app are named; these are useful because the same name will often be referenced by both the server and the ui portions, so naming conventions make it so you don't have to remember as many variable names; these are also useful because similar variables will often appear for each of the datatypes (clinical, assay, feature, all) and naming conventions make it so you can name these all similarly and reference each of them without having to remember how you named them
    + `time_testing.R`
        - basically a sandbox for optimizing various pieces of code; se the [time testing tips](#time-testing) section
    + `VERSION`
        - defines the version of the app, updates with every commit
    + `help_docs/`
        - Markdown files for the help icons in the app
        - (Generally it's a good idea to separate out large portions of text from code that actually does work.)
    + `rsconnect/`
        - an automatically-generated directory that I do not touch; necessary to publish the app to shinyapps.io
    + `series_platform/`
        - contains a script that populates the dropdown you see on the first tab
    + `server/`
        - contains all the code for the server function
        - subsectioned by datatype (clinical, assay, feature, all)
    + `ui/`
        - contains all the code for the ui object
        - subsectioned by datatype (clinical, assay, feature, all)
    + `User/`
        - pretty much everything in this directory is useless except rscript_functions.rds--it is the list of the               functions that can be included in the script the user can download
    + `www/`
        - resources for the app, including help gifs, the series and platform lists, the javascript file, and the css           styling file
* Level 2 - `server`
    + Each directory contains R scripts organized by tab (the name of the tab is usually the name of the R script, and       the tab is named by what the user does in that tab)
    + `assay/`
        - `feature_data.R` is the code for the modal that comes up when you click the "Replace ID" button
        - `side_panel.R` and `helper_functions.R` are pretty much useless (the code in helper_functions.R should all be in `formatting_functions.R` now)
    + `clinical/`
        - `helper_functions.R` is pretty much useless (the code in `helper_functions.R` should all be in `formatting_functions.R` now)
    + `formatting_helper_functions.R`
        - the functions that can be included in the script the user can download; __these same functions__ are used in the app to format the data in various ways--usually one function per tab
    + `help_modals.R`
        - functions to create the modals that come up when the user clicks a question icon, observers to make the               question icons reactive
    + `regex_modal.R`
        - the code to create the modal that allows users to test regular expressions
    + `tidygeo_server_functions.R`
        - functions to create various pieces of the server; these often correspond to similarly-named ui                        functions to create an object the user can interact with
* Level 2 - `ui`
    + Each directory contains R scripts organized by tab (the name of the tab is usually the name of the R script, and       the tab is named by what the user does in that tab)
    + `assay/`
        - `side_panel.R` and `side_panel_assay.R` are useless
    + `feature/`
        - `side_panel_feature.R` is useless
    + `tidygeo_ui_functions.R`
        - functions to create various pieces of the ui; these often correspond to similarly-named server                        functions to create an object the user can interact with
* Level 2 - `www`
    + `reactive_preferences.js`
        - when I have a bunch of buttons that do the same thing, I can't name the buttons the same thing (duplicate             IDs not allowed) and I can't have an `observeEvent` that listens to multiple buttons; instead, I give the               buttons a class and whenever a button from that class is clicked, the JavaScript here puts that button's              ID in an input for me to listen to with an `observeEvent`; after I've used that input, I have to reset it               with this JavaScript so the next button click can set off the `observeEvent` again
        - it's generally good practice to put any JS in a separate script such as this one so that your R code                  doesn't look as cluttered
    + `style.css`
        - a bunch of HTML that defines how certain elements of the app should look
        - for each chunk, you'll see a dot followed by the name of a class followed by some attributes for the class
        - it's generally good practice to put any styling HTML in a separate script such as this one so that your R code doesn't look as cluttered

### Style guide and naming practices {#style-guide}
I tried my best to follow the tidyverse style guide written by Hadley Wickham as it appears in [this article](https://style.tidyverse.org/).

#### Naming conventions
1. Most variable and function names are all undercase with underscores separating the words. If you ever want to change this convention, please change the `SEP` variable in tidygeo_variables.R
2. Much of the app is built using functions, and these functions use specific names to reference certain variable names in the app. Rather than having the programmer memorize how the functions reference the variable names, I created naming patterns in `tidygeo_variables.R` that define how certain types of variables are named. To change the naming pattern, one simply changes `tidygeo_variables.R` and all the variable names in the app will be updated. In this way, the naming conventions in `tidygeo_variables.R` also prevent extensive refactoring.
3. In general, variables are named using three parts: the general name for the variable, the datatype the variable corresponds to (clinical, assay, feature, all), and some kind of extra tag to make that variable unique, in case it is used for the same datatype in multiple sections.

#### Organization
1. Tabs are organized by datatype and by what the user can do to that data in the tab. 
2. Code is usually organized by datatype and tab. Code chunks are separated by a comment ending in three or more dashes, which can be easily produced in RStudio using `Ctrl` + `Shift` + `R`.
3. Chunks of text for the app are usually kept in separate files, not in the R code.
4. All global variables are defined in `tidygeo_variables.R`, not in app.R or any of the scripts that build the app.
5. All libraries needed for the app to run are loaded in `tidygeo_variables.R`.
6. Functions are always defined outside the tab R scripts. Functions that are used in the UI portion are defined in `tidygeo_ui_functions.R`. Functions that make changes to the data are defined in `formatting_functions.R`. Functions that are used in the Server portion are defined in `tidygeo_server_functions.R`. Functions that do not fit into a category are defined in `tidygeo_functions.R`.

### Functions and how to use them {#tidygeo-functions}
There are two reasons that I would __strongly encourage__ any programmer working on this app to use the existing functions rather than writing things from scratch.

1. Pieces of UI often correspond to similarly-named pieces of Server.

Just like you pair a textOutput in the UI with a renderText in the Server, I have created pieces of UI that correspond to pieces of the Server, and these corresponding pieces have similar names. One example is with the column navigation you might notice above each of the datatables in the app. The UI for the column navigation looks like this:
```{r}
#ui
```
and the server looks like this:
```{r}
#server
```
See how the variable names in the UI and Server look similar? Now, rather than repeating these chunks of UI and Server for every single datatype, I made functions that will create these pieces for me:
```{r}
# UI functions
# server functions
# ui
# server
```
2. The use of functions makes variable names very important

You might have noticed how the functions above had to know a little bit about how the variables would be named in order to

3. Functions have error-checking in place

[&uarr; To top](#contents)

------------------------

## Next Steps {#next-steps}

### Known bugs {#known-bugs}

### TODO items {#todo-items}

### Possible future features {#future-features}

[&uarr; To top](#contents)

------------------------

## Miscellaneous {#misc}

### Error log {#error-log}
When I run into a tricky error, I make note of it in `error_log.md`, so when I encounter the error again, I can know how to fix it quickly. Here are some of the errors I have encountered so far.

------------------------

```{r child = 'TidyGEO/error_log.md'}
```

------------------------

### Shiny/R tips and observations {#observations}

#### On renderUI vs updateInput
`renderUI`s are, in general, better than observes with `update[]Input`s in them. This is because observes are called more often--whenever their dependencies are updated--rather than only when we need to use them in the UI. I have a hypothesis, though, that with computationally-intensive UIs that we need to use often and that update often, an `update[]Input` might outperform a `renderUI` since it only has to update a part of the UI rather than render the entire thing again. However, I have never encountered a UI so difficult to render that `renderUI` lagged down the code. So IDK.

### Tips for time testing {#time-testing}

### Other resources {#other-resources}

#### On Rmarkdown {#Rmarkdown-resource}
Pimp my RMD: a few tips for R Markdown, by Yan Holtz [(link)](https://holtzy.github.io/Pimp-my-rmd/)

[&uarr; To top](#contents)
